Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2022-03-22 doc/peakcharacterization_manuscript.Rmd
incorporating Joshs edits

To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as doing a straight averaging of point intensities using identical scans as was kept by our scan filtering (see Methods) in the xcms R package, and finally averaging peaks that were centroided via the Thermo raw file parsing utilities (wrapped by ThermoRawFileParser). 
In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks are off compared to what is expected based on natural abundance probability (NAP), as shown in Figure X, with some solutions giving different amounts of error. 
We do note that when all of the peaks are present in a single scan, their relative intensity ratios are much closer to the theoretical ratios based on NAP.

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. 
Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans. 
However, the frequency can be calculated by dividing the average m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the average m/z of two adjacent points, red lines representing the difference between the two adjacent points). 
The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C. 
The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D. 
Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E. 
This model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$. 
This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
The m/z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the lasso model.
Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.

@@

In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis. 
It is expected that noise is randomly distributed across the scans. 
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which we think any non-zero points are potentially **signal** and not just noise.
In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan. 
By discarding those windows that have a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.

@@

`r figure_count$label_text("slidingwindow_count")`. 
**A**: Histogram of the number of non-zero points from sliding windows 10 frequency points wide and 1 frequency point apart, from a single tiled window 2000 points wide.
Blue line is the 99th percentile.
**B**: Histogram of the 99th percentile cutoffs from all of the tiled windows.
Red line is the median x 1.5.
**C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.

@@

Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly (see `r figure_count$label_text("breaking_regions")`A and B for example).
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan order; and 2) some peak differences between scans are correlated with intensity.
The solution to **1** is to do a two pass normalization.
After the first pass, the peaks who's intensity is correlated with scan order are detected.
In the second pass, the correlated peaks are removed, and normalization is carried out again.
`r figure_count$label_text("intensity_scan")`A contains an example peak whose intensity across scans is correlated with scan number.
The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.

@@

We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018]. 
Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
Their presence should be minimized by removing noise peaks, and removing regions that have multiple reported peaks in the same scan. 
We expect they may present as characterized peaks that have larger than expected frequency standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
To verify the removal of HPD peaks, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
Peak density was measured using a sliding window 10 points wide and offset by one point.

@@

The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan. 
There is no information about the original frequency values available in either the `raw` files or the `mzML` files. 
However, the frequency values can be approximated by averaging the m/z of adjacent points, and dividing the m/z by the difference. 
Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value. 
These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:

$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
From the known physical properties of the Orbitrap, only the square root term should be necessary.
Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer. 
A frequency model is generated for each scan, and then a single model using the scan with the square-root slope closest to the median of the square-root slopes from all scans.
The single model is used to convert the m/z values to frequency for all scans.
A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).

@@

Two types of frequency intervals are used, sliding and tiled windows. 
In this work, the sliding windows are 10 points wide and offset by one point.
The tiled windows are one point wide and offset by one point.

@@

To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a]. 
The sliding and tiled windows are also converted to IRanges objects the same way.

@@

To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018]. **aside: I'm not sure how this actually relates to the density noise cutoff**.

@@

Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks. 
In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans. 
The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan. 
This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.

@@

Based on the previously detected peaks, the full set of raw data points for each peak in each scan within a region is known. 
The non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be derived from the scan level peak characteristics previously calculated.

@@

Ideally, each peak would be observed in every scan. 
However, in some scans, some peaks are not likely to be observed due to the number of ions falling below the detection threshold. 
This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans. 
To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used. 
For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated. 
The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022].

@@

High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations. Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space. 
Sliding windows that are 1000 frequency points wide and offset by 100 points were used for the density calculations.

@@

To evaluate the F-statistic between multi-class samples, a much larger set of samples was used. In this case, 181 matched non-cancer and cancer samples that were previously used in the HPD detection manuscript [ref]. All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the xcms R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF). For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from xcms.

@@

Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference]. 
All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode) and ThermoRawFileParser (centroided).

@@

  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
  1. Scan level peak centroids generated by MSnbase
  1. Scans merged and then centroids generated by MSnbase

>>>>

To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as merging scans (see Methods) in the MSnbase R package. 
In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks deviate from the intensity ratio predicted by their natural abundance probabilities (NAPs), as shown in Figure X, with some solutions giving different amounts of error. 
We do note that when all of the peaks are present in a single scan, their relative intensity ratios are much closer to the theoretical ratios based on NAP.

@@

FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw frequency data. 
Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans. 
However, the frequency can be calculated by dividing the average m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the average m/z of two adjacent points, red lines representing the difference between the two adjacent points). 
The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C. 
The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D. 
Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E. 
This model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$. 
This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
The m/z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the lasso model.
Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.

@@

In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis. 
Furthermore, it is expected that noise is randomly distributed across the scans. 
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which any non-zero points are potentially **signal** and not just noise.
In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan. 
By discarding those windows with a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.

@@

`r figure_count$label_text("slidingwindow_count")`. 
**A**: Histogram of the number of non-zero points from sliding windows ten frequency points wide and a stride of one point, from a single tiled window 2000 points wide.
Blue line is the 99th percentile.
**B**: Histogram of the 99th percentile cutoffs from all of the tiled windows.
Red line is the median x 1.5.
**C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.

@@

Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary slightly between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan number (i.e. scan acquisition order); and 2) some peak differences between scans are correlated with intensity.
The solution to **1** is to do a two pass normalization.
After the first pass, the peaks whose intensity is correlated with scan order are detected.
In the second pass, the correlated peaks are removed, and normalization is carried out again.
`r figure_count$label_text("intensity_scan")`A contains an example peak whose intensity across scans is correlated with scan number.
The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.

@@

We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018]. 
Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan. 
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and removing those that are outliers.
To verify the removal of HPD peaks, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
Peak density was measured using a sliding window ten points wide and a stride of one point.

@@

The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan. 
No information about the original observed frequency values is available in either the `raw` files or the `mzML` files. 
However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the difference. 
Ideally, the difference between subsequent points in this proxy frequency space is constant, in practice there is a range of differences in frequency space.
The *actual, constant* difference can be obtained by examining the median of the frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value. 
These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:

$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
A frequency model is generated for each scan, and then a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.
We observed that this single model better preserved the relative ordering of the peaks in both m/z and frequency-space compared to the scan specific models (see Results).
A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).

@@

Two types of frequency intervals are used, sliding and tiled windows. 
In this work, the sliding windows are 10 points wide with a stride of one point.
The tiled windows are one point wide with a stride of one point.

@@

To enable interval algebra, the frequency points are converted to single width intervals by multiplying (by a multiplier of 400 in this work), rounding to the nearest integer, and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a]. 
The sliding and tiled windows are also converted to IRanges objects using this process.

@@

To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged to create the initial peak regions.
The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].

@@

Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks present in at least the same number of scans as the 95th percentile of scan counts for the peaks. 
In addition, only those peaks with an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans. 
The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan. 
This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.

@@

The full set of raw data points for each peak in each scan within a region is known based on the previously detected peaks. 
Therefore, the non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
In addition to the data from the full set of raw points, the means and the standard deviations of the peak height and location can be derived from the scan-level peak characteristics previously calculated.

@@

Ideally, each peak would be observed in every scan. 
However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold. 
This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans. 
To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used. 
For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated. 
The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022; @burkardtTruncatedNormalDistribution].

@@

High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations. 
Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space. Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.

@@

To evaluate the F-statistic between multi-class samples, a much larger set of samples was used. In this case, 181 matched non-cancer and cancer samples previously used in the HPD detection manuscript [@mitchellNewMethodsIdentify2018]. 
All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the MSnbase R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF). For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using scan-level peak-characterization and those from MSnbase.

@@

Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference]. 
All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode).

@@

  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
  1. Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
  1. Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
